---
title: "air quality circulant"
output: html_document
date: "2025-05-06"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rootSolve)
```

### Circulant model estimation
 - Circulant model by creating a 24 dimensional model via the hourly observations.
 - Treat each day as an independent realization
```{r}
folder <- 'Data/Beijing_data/PRSA_Data_20130301-20170228/'
files <- list.files(folder)
df_list = lapply(paste0(folder,files), read_csv)
df1 <- df_list[[6]]
```


### ACF alternative
```{r}
log(df1$PM10) %>% na.omit %>% acf(main = '')
```

#### Gaussian kernel and exponential covariance function fitted to ACF
```{r}
acf_obj <- log(df1$PM10) %>% na.omit %>% acf(plot = F)

lags <- as.numeric(acf_obj$lag)
acf_vals <- as.numeric(acf_obj$acf)

df_acf <- data.frame(lag = lags, rho = acf_vals)

fit_gauss <- nls(rho ~ exp(-lag^2 / (2 * sigma^2)),
           data = df_acf,
           start = list(sigma = 5),
           control = nls.control(maxiter = 100))

fit_exp <- nls(rho ~ exp(-a*lag),
           data = df_acf,
           start = list(a = 1),
           control = nls.control(maxiter = 100))

plot(acf_obj, main = "")
lines(df_acf$lag,
      predict(fit_gauss),
      col = "red", lty = 'dashed', lwd = 1.5)
lines(df_acf$lag,
      predict(fit_exp),
      col = "orange", lty = 'dashed', lwd = 1.5)

legend("topright", legend = c("Gaussian fit", "Exponential fit"), 
       col = c("red", "orange"), 
       lwd = c(1.5, 1.5), lty = c('dashed', 'dashed'))

# AIC(fit_gauss, fit_exp)
```




### Periodogram

```{r}
spec <- stats::spec.pgram(
  log(df1$PM10) %>% na.omit,
  taper = 0,
  plot = FALSE,
  detrend = F,
  demean = T
)


periods <- 1 / spec$freq

plot(
  periods, spec$spec,
  type = "l",        
  xlim = c(0, 500),  # up up to 500-hour cycles
  xlab = "Period (hours)",
  ylab = "Spectral density",
  main = "Periodogram of hourly PM10", 
  lwd = 1
)
abline(v = c(24, 168), col = rgb(255, 0, 0, 0.4*255, maxColorValue = 255), lty = 'dashed', lwd = 1)
text(c(24,168), 
     par("usr")[4]*c(0.8,0.8), 
     labels = c("24h (1 day)","168h (1 week)"), 
     col="red", adj = c(0,1))
```

```{r}
s <- spec$spec

# local maxima where s[i] > s[i-1] and s[i] > s[i+1]
peak_idx <- which(diff(sign(diff(s))) == -2) + 1

peak_periods <- periods[peak_idx]
peak_power   <- s[peak_idx]


top20 <- order(peak_power, decreasing=TRUE)[1:20]

data.frame(
  period = peak_periods[top20],
  power  = peak_power[top20]
)
```


### Mean per hour and weekday/weekend

```{r}
global_mean <- df1 %>% 
  summarise(m = mean(log(PM10), na.rm=TRUE)) %>% 
  pull(m)

df_summary <- df1 %>%
  mutate(
    log_obs = log(PM10),
    dt = make_datetime(year, month, day, hour, tz = "Asia/Shanghai"),
    weekday = wday(dt, label=TRUE, abbr=FALSE, locale = 'en_US'),
    Weekend = factor(weekday %in% c("Saturday","Sunday"), levels = c(FALSE,TRUE), 
                     labels = c("Weekday","Weekend"))
  ) %>% 
  group_by(hour, Weekend) %>%
  summarise(mean_hour = mean(log_obs, na.rm=TRUE), .groups="drop")

ggplot(df_summary, aes(hour+1, mean_hour)) +
  geom_line(aes(linetype = Weekend)) +
  geom_hline(aes(yintercept = global_mean,
                 linetype     = "Global mean"),
             color = "red", alpha = 0.8) +
  scale_linetype_manual(
    name   = NULL,
    values = c(
      "Weekday"     = "solid",
      "Weekend"     = "dotted",
      "Global mean" = "dashed"
    ),
    guide = guide_legend(
      override.aes = list(
        color = c("black","black","red")
      )
    )
  ) +
  labs(x = "Hour of day", y = "Mean log(PM10)") +
  scale_x_continuous(
    breaks = seq(1, 24, by = 2),
    labels = function(x) sprintf("%02d:00", x)
  ) +
  theme_minimal()

```

### Data wrangling to get a data matrix
m = days
dim af data: days x 24 med en af de målte variable

```{r}
df_ <- df1 %>% 
  mutate(date = make_date(year, month, day)) %>% 
  mutate(log_obs = log(PM10)) %>% 
  select(date, hour, log_obs)
  

df_wide <- df_ %>% pivot_wider(id_cols = date,
                    names_from = hour, 
                    values_from = log_obs,
                    names_prefix = 'h') %>% na.omit
```

```{r}
data1 <- df_wide %>% 
  select(-date) %>% 
  as.matrix %>% unname
```




### Numerical estimation 
```{r}
R_val <- function(v, mu, k) {
  N <- length(v)
  v_ = v- mu
  shifted_indices <- ((0:(N-1) + k ) %% N) + 1
  return( sum(v_ * v_[shifted_indices]) ) 
}

R_val_vec <- Vectorize(R_val, 'k')


kappa_hat_numerical <- function(beta_vec, mu, V) {
  m = nrow(V)
  N = ncol(V)
  p = length(beta_vec)
  sum_squares <- sum(apply(V, 1, function(x) sum((x-mu)^2)))
  R_vals = apply(V, 1, function(x) R_val_vec(x, mu, 1:p))
  denom = sum_squares + 2*sum(beta_vec*R_vals)
  return(N*m/denom)
}

score_beta_numerical <- function(beta_vec, V, mu) {
  N <- ncol(V)
  m <- nrow(V)
  p <- length(beta_vec)
  f <- numeric(p)
  
  kappa_est = kappa_hat_numerical(beta_vec, mu, V)
  
  for (k in 1:p) {
    
    j_seq = 0:(N-1)
    
    term1 <- m*sum(sapply(j_seq, function(j) {
      numerator = cos(2*pi*j*k/N)
      cosines_denom_j = cos(2*pi*j*(1:p)/N)
      denom = 1 + 2*sum(beta_vec*cosines_denom_j)
      
      return(numerator/denom)
    }))
    
    R_val_k = sum(apply(V, 1, function(x) R_val(x, mu, k)))
    term2 = kappa_est*R_val_k
    
    f[k] <- term1 - term2
  }
  return(f)
}

score_beta_numerical_single <- function(beta_vec, V, mu) {
  N <- ncol(V)
  m <- nrow(V)
  p <- length(beta_vec)
  f <- numeric(p)
  
  kappa_est = kappa_hat_numerical(beta_vec, mu, V)
  
  for (k in 1:p) {
    
    j_seq = 0:(N-1)
    
    term1 <- m*sum(sapply(j_seq, function(j) {
      numerator = cos(2*pi*j*k/N)
      cosines_denom_j = cos(2*pi*j*(1:p)/N)
      denom = 1 + 2*sum(beta_vec*cosines_denom_j)
      
      return(numerator/denom)
    }))
    
    R_val_k = sum(apply(V, 1, function(x) R_val(x, mu, k)))
    term2 = kappa_est*R_val_k
    
    f[k] <- term1 - term2
  }
  return(f)
}

solve_params_numerical <- function(initial_guess_betas, V) {
  mu_est <- mean(V)
  solution <- multiroot(f = function(bet) score_beta_numerical(bet, V, mu_est), start = initial_guess_betas)
  beta_est_num <- solution$root
  kappa_est_num = kappa_hat_numerical(beta_est_num, mu_est, V)
  
  return(
    list('kappa_est' = kappa_est_num,
         'beta_est' = beta_est_num,
         'mu_est' = mu_est)
  )
}
solve_params_numerical_single <- function(V) {
  mu_est <- mean(V)
  solution <- uniroot(f = function(bet) score_beta_numerical(bet, V, mu_est), lower = -0.5, upper = 0.5)
  beta_est_num <- solution$root
  kappa_est_num = kappa_hat_numerical(beta_est_num, mu_est, V)
  
  return(
    list('kappa_est' = kappa_est_num,
         'beta_est' = beta_est_num,
         'mu_est' = mu_est)
  )
}
```

### Analytical estimation
```{r}
# A function to cyclically shift a vector by 'shift_by' positions to the left
cyclic_shift <- function(vec, shift_by) {
  n <- length(vec)
  if (shift_by == 0) {
    return(vec)
  }
  # shift by 1: [2:n, 1].  
  # shift by 2: [3:n, 1:2].
  return(c(vec[(shift_by+1):n], vec[1:shift_by]))
}


# Cycle-average covariance from ONE vector of length N
cov_cycle_one <- function(x) {
  N <- length(x)
  j_seq <- 0:(N-1)
  outer_prods <- lapply(j_seq, function(j) {
    x_shift <- cyclic_shift(x, j)
    return(outer(x_shift, x_shift))
  })
  return(Reduce("+", outer_prods) / N)
}


# Handle M realizations
cov_cycle_multi <- function(V, mu_est) {
  V = V-mu_est
  # V is m x N
  m <- nrow(V)
  N <- ncol(V)
  
  cov_list <- vector("list", m)
  for (i in seq_len(m)) {
    x_i <- V[i, ]  # one realization
    cov_list[[i]] <- cov_cycle_one(x_i)
  }
  Sigma_est <- Reduce("+", cov_list) / m  
  return(Sigma_est)
}

solve_params_analytical <- function(V, p) {
  mu_est = mean(V)
  Sigma_est <- cov_cycle_multi(V, mu_est)
  Q_est <- solve(Sigma_est)
  kappa_est <- Q_est[1,1]  

  beta_est <- numeric(p)
  for (k in 1:p) {
    Q_0k <- Q_est[1, 1+k]
    beta_est[k] <- Q_0k / kappa_est
  }
  
  return(
    list('kappa_est' = kappa_est,
         'beta_est' = beta_est)
  )
}
```


## Estimates on all the data
```{r}
analytical_ests <- solve_params_analytical(data1, 11)
est_num_single <- solve_params_numerical_single(data1)
#solve_params_numerical(analytical_ests$beta_est, data1)

```

```{r}
plot(1:11, analytical_ests$beta_est, type = 'h', lwd = 1.5, ylab = '', xaxt = 'n', xlab = '')
points(analytical_ests$beta_est, pch = 20)
abline(h = 0, lty = 'dashed')
axis(
  side = 1,
  at = 1:11,
  labels = parse(text = paste0("beta[", 1:11, "]")),
  las = 1
)
```

 
 
## Residuals and partial correlation of these
```{r}

mu_est  <- est_num_single$mu_est
beta_est <- est_num_single$beta_est  
N   <- ncol(data1)

fitted_mat <- t( apply(data1, 1, function(v){
  im1   <- c(N, 1:(N-1))   
  ip1   <- c(2:N, 1)       
  mu_est - beta_est*((v[im1]-mu_est)+(v[ip1]-mu_est))
}) )

R_in <- data1 - fitted_mat   # residuals matrix

S  <- cov(R_in)

prec_R <- solve(S)

pcor_mat <- -prec_R / sqrt( outer(diag(prec_R), diag(prec_R)) )
diag(pcor_mat) <- 1

# pcor_mat[i,j] is the partial correlation between residuals
# at hour i and hour j, controlling for all other hours.


```

 
```{r}
h12_pcor = pcor_mat[12,]
h12_pcor[12] = NA
plot(1:24, h12_pcor, xlab = 'Hour', ylab = 'Conditional corr. with H12', pch = 20)
```

 
## Estimates for each weekday
```{r}
Sys.setlocale('LC_TIME', 'en_US')
create_weekday_data_matrix <- function(df, dow) {
  df_wday = df %>% 
    mutate(date = make_date(year, month, day),
           weekday = weekdays(date),
           log_obs = log(PM10)) %>% 
    filter(weekday == dow) %>% 
    select(date, hour, log_obs)
  
  df_wider <- df_wday %>% pivot_wider(id_cols = date,
                    names_from = hour, 
                    values_from = log_obs,
                    names_prefix = 'h') %>% na.omit
  
  data_mat <- df_wider %>% 
    select(-date) %>% 
    as.matrix %>% unname
  return(data_mat)
}

dows = c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 
         'Saturday', 'Sunday')


result_list = list()
mus = numeric(7)
names(mus) = dows

kappas = numeric(7)
names(kappas) = dows

betas = numeric(7)
names(betas) = dows

for (dow in dows) {
  dat = create_weekday_data_matrix(df1, dow)
  ests_analytical = solve_params_analytical(dat, 12)
  numerical_est_single = solve_params_numerical_single(dat)
  result_list[[dow]] = unlist(numerical_est_single)
  mus[dow] = numerical_est_single$mu_est
  kappas[dow] = numerical_est_single$kappa_est
  betas[dow] = numerical_est_single$beta_est
}
```




## Crossvalidation out of sample predictions

Forecasting - brug kun den forrige nabo


Vælg en enkelt dag og brug alle 24 timer som eksempel
```{r}
days <- 1:nrow(data1)
errors <- numeric(nrow(data1))
preds <- numeric(nrow(data1))
true_vals <- numeric(nrow(data1))
hours_left_out = numeric(nrow(data1))

for (d in days) {
  set.seed(42 + d)
  train_idx = setdiff(days, d)
  data_train = data1[train_idx,]
  data_test = data1[d,]
  
  params <- solve_params_analytical(data_train, 1)
  mu_cv = mean(data_train)
  beta_cv <- params$beta_est
  
  
  h <- sample(1:24, 1)
  hours_left_out[d] <- h
  
  hm1 = if(h==1) 24 else h-1
  hp1 = if(h==24) 1 else h+1
  
  Y_obs <- data_test
  true_vals[d] <- Y_obs[h]

  Y_obs[h] <- NA
  pred_h <- mu_cv - beta_cv*((Y_obs[hm1]-mu_cv)+(Y_obs[hp1]-mu_cv))
  preds[d] <- pred_h
  
  errors[d] <- (pred_h - data_test[h])^2
}


#single day example
d = 200
train_idx = setdiff(days, d)
data_train = data1[train_idx,]
data_test = data1[d,]

params <- solve_params_analytical(data_train, 1)
mu_cv = mean(data_train)
beta_cv <- params$beta_est[1]

true_vals_single_day = numeric(24)
preds_single_day = numeric(24)

for (h in 1:24) {
  
  hm1 = if(h==1) 24 else h-1
  hp1 = if(h==24) 1 else h+1
  
  Y_obs <- data_test
  true_vals_single_day[h] <- Y_obs[h]
  
  Y_obs[h] <- NA
  pred_h <- mu_cv - beta_cv*((Y_obs[hm1]-mu_cv)+(Y_obs[hp1]-mu_cv))
  preds_single_day[h] <- pred_h
}
```


```{r}
plot(1:24, true_vals_single_day, type = 'o', xlab = 'Hour', ylab = 'log(PM10)')
points(1:24, preds_single_day, col = 'red', pch = 4, type = 'o')
legend('topleft', legend = c('Predicted', 'True'), pch = c(4, 1), col = c('red', 'black'), cex = 0.9, bty = 'n')
```


## Errors grouped on hours
```{r}
df_errors <- data.frame(hours_left_out, preds, true_vals)

errors_per_hour <- df_errors %>% 
  group_by(hours_left_out) %>% 
  summarise(error = sqrt(mean((preds-true_vals)^2)),
            n = n())

errors_per_hour %>% 
  ggplot(aes(x = hours_left_out, y = error)) + 
  geom_col() +
  labs(x = "Hour of day", y = "RMSE") +
  scale_x_continuous(
    breaks = seq(1, 24, by = 2)
  ) +
  geom_hline(aes(yintercept = sqrt(mean(errors))), linetype = 'dashed', col = 'red', alpha = 0.8)+
  theme_minimal()
```

