---
title: "Spatial og cirkulant kombination"
output: html_document
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Gaussian kernel function for points evenly spaced on a circle


```{r}
index_diff <- function(i,j, N) {
  return(
    pmin((i-j) %% N, (j-i) %% N)
  )
}

# Distances based on arc lengths
gauss_kernel_circ <- function(mu, sig_kernel, N) {
  D = outer(1:N, 1:N, function(i,j) index_diff(i,j, N)*2*pi/N)
  D_sq = D^2
  mat = exp(-D_sq/(2*sig_kernel^2))
  diag(mat) = 0
  return(mat)
}

```




## Functions to simulate from p'th order circulant GMRF
```{r}
circulant_matrix_base_p = function(kappa, betas, N) {
  order = length(betas)
  diag_entry = kappa
  remainder = c(kappa*betas, 
                rep(0, N-1-2*order), 
                rev(kappa*betas))
  return( c(diag_entry, remainder) )
}

create_circulant_precision_matrix_p = function(kappa, betas, N = 10) {
  base = circulant_matrix_base_p(kappa, betas, N)
  N = length(base)
  idx = outer(seq_len(N), seq_len(N), function(i,j) ((j-i) %% N) +1)  # Definition 2.4 Rue & Held
  matrix_out = matrix(base[idx], nrow = N)
  return (matrix_out)
}

simulate_circulant_GMRF = function(mu, precision_mat, nsim = 1) {
  Q = precision_mat
  N = nrow(Q)
  Sigma = solve(Q) 
  out = MASS::mvrnorm(n = nsim, mu = rep(mu, N), Sigma = Sigma)
  return(out)
}

```



## Estimating circulant model parameters via kernel model specification

- We simulate from a circulant model 
- Recover kernel based model parameters from simulated data
- Convert kernel model parameter estimates to circulant model parameter estimates

### kernel model log-likelihood
```{r}
log_like_kernel_single <- function(V_vec, W, W_tilde, eigenvals, rowsums, mu, phi, lambda) {
  N = length(V_vec)
  v_ = V_vec-mu 
  
  
  term1 = -N/2*log(lambda)
  term2 = 1/2*sum(log(rowsums))
  term3 = 1/2*sum(log(1 - phi*eigenvals))
  term4 = -N/2*log(2*pi)
  term5 = -1/(2*lambda)*sum(v_^2*rowsums)
  term6 = phi/(2*lambda)*(v_%*%W%*%v_)
  
  return(term1 + term2 + term3 + term4 + term5 + term6)
}

log_like_kernel <- function(V, W, W_tilde, eigenvals, rowsums, mu, phi, lambda){
  apply(V, 1, function(x) log_like_kernel_single(x, W, W_tilde, eigenvals, rowsums, mu, phi, lambda)) %>% 
    sum %>% return
}
```

### Parameter estimation functions for kernel model
```{r}
# mu_est <- mean(V)


quad_terms_single_v <- function(v, mu, phi, W, rowsums) { 
  v <- v - mu
  term1 <- sum(v^2*rowsums)
  term2 <- phi* (v%*%W%*%v)
  return(term1 - term2)
}

lam_estimate <- function(v, mu, phi, W, rowsums) {
  m <- nrow(v)
  N <- ncol(v)
  quad_terms <- apply(v, 1, function(v) quad_terms_single_v(v, mu, phi, W, rowsums))
  return (sum(quad_terms) / (N*m))
}

# eigenvalues of W_tilde - calculate once 
score_phi <- function(phi, v, mu, eigenvals_W_tilde, W, rowsums) {
  N <- ncol(v)
  m <- nrow(v)
  term1 <- -m/2 * sum( eigenvals_W_tilde/(1-phi*eigenvals_W_tilde))
  
  sum_quads_denom = sum(apply(v, 1, function(v_) quad_terms_single_v(v_, mu, phi, W, rowsums)))
  sum_quads_numer = sum(apply(v, 1, function(v_) {
    v = v_-mu
    return(v%*%W%*%v)
  }))
  term2 <- N*m/(2*sum_quads_denom) *sum_quads_numer
  return(term1 + term2)
}

# res <- uniroot(
#   f = function(phi) score_phi(phi, V, mu_est, lambdas, W, rowsums),
#   lower = -1,
#   upper = 0.99
# )
# 
# phi_est <- res$root
# lam_est <- lam_estimate(V, mu_est, phi_est, W, rowsums)
# 
# c('phi_est'=phi_est, 'lambda_est' = lam_est, 'mu_est' = mu_est)
```


- Regn likelihood for parametre estimeret givet et bestemt sig_kernel.
### Varying sigma (for kernel), calculating log like for each and choosing MLE
```{r}
sigmas_kernel = seq(0.01, 1, by = 0.01)
n_sigmas = length(sigmas_kernel)

N <- 20
set.seed(123)
mu_true = 0
kappa_true = 10
betas = c(-0.18, -0.12, -0.06, -0.03)
Q = create_circulant_precision_matrix_p(kappa_true, betas, N = N)
V = simulate_circulant_GMRF(mu_true, Q, nsim = 500)

mu_est = mean(V)

lls = numeric(length(sigmas_kernel))

for (i in 1:n_sigmas) {
  sigma = sigmas_kernel[i]
  
  W = gauss_kernel_circ(mu_est, sigma, N)
  rowsums <- rowSums(W)
  c <- rowsums[1]
  
  W_tilde = (1 / c) * W
  lambdas <- eigen(W_tilde)$values
  
  res <- uniroot(
      f = function(phi)
        score_phi(phi, V, mu_est, lambdas, W, rowsums),
      lower = -1,
      upper = 0.99
    )
  phi_est <- res$root
  lam_est <- lam_estimate(V, mu_est, phi_est, W, rowsums)
  ll = log_like_kernel(V, W, W_tilde, lambdas, rowsums, mu_est, phi_est, lam_est)
  
  lls[i] <- ll
}


## endelige estimater ud fra max likelihood 
sigma_max_ll = sigmas_kernel[which.max(lls)]

W = gauss_kernel_circ(mu_est, sigma_max_ll, N)
rowsums <- rowSums(W)
c <- rowsums[1]

W_tilde = (1 / c) * W
lambdas <- eigen(W_tilde)$values

res <- uniroot(
    f = function(phi)
      score_phi(phi, V, mu_est, lambdas, W, rowsums),
    lower = -1,
    upper = 0.99
  )
phi_est <- res$root
lam_est <- lam_estimate(V, mu_est, phi_est, W, rowsums)


```

```{r}
plot(sigmas_kernel, lls, type = 'l', ylab = 'log-likelihood', xlab = expression(sigma))
abline(v = sigma_max_ll, lty = 'dashed', col = 'red')
legend('topright', legend = expression(sigma['*']),
       lty = 'dashed', col = 'red', bty = 'n')
```


Konvertering til cirkulant model estimater.
```{r}
kappa_est <- c/lam_est

beta_est <- numeric(N/2)
for (k in 1:(N/2)) {
  beta_est[k] = -(phi_est/c)* W[1,k+1]  # samme
}
circ_kernel = gauss_kernel_circ(5, sig_kernel = sigma_max_ll, N)[1,2:(N/2 +1)]
```


```{r}
# define a fine grid of k’s from 1 to N
k_grid1 <- seq(1, N/2, length.out = 500) # only need this part later
k_grid2 <- seq(N/2, N, length.out = 500)
k_grid = c(k_grid1, k_grid2)

# map to angles in [0,2*pi]
theta_grid <- (k_grid) * 2 * pi / N

# circular distance from 0
d_grid <- pmin(theta_grid, 2*pi - theta_grid)

# continuous Gaussian kernel scaled by phi_est/c
kernel_cont <- (phi_est/c) * exp(-d_grid^2 / (2 * sigma_max_ll^2))

```


Plot with true betas and estimated betas from kernel
```{r}
# beta_true <- c(beta1_true, beta2_true, beta3_true, rep(0, N/2 - 3))
# beta_true <- c(beta1_true, beta2_true, rep(0, N/2 - 2))
# beta_true <- c(beta1_true, beta2_true, beta3_true, beta4_true, rep(0, N/2 - 4))
beta_true <- c(betas, rep(0, N/2 - length(betas)))

plot(-beta_est, 
     ylim = c(min(0, min(-beta_est)), max(max(-beta_est), -beta_true)),
     pch = 3, 
     lwd = 1.5,
     xlab = 'k', 
     ylab = latex2exp::TeX(r'($-\beta_k$)'))

points(-beta_true, col = 'red', pch = 2, lwd = 1.5)
lines(k_grid1, kernel_cont[1:500], lty = 'dashed')  # continous interpolation

legend("topright", 
       legend = c("Estimate", "True value", "Gaussian kernel \n(scaled appropriately)"),
       col = c("black", "red", "black"),
       pch = c(3, 2, NA),
       lty = c(NA, NA, "dashed"),
       lwd = c(1.5, 1.5, 1.5),
       bty = "n",
       cex = 0.9)
```




## Kernel model with relation to circulant model


Functions to simulate from kernel model GMRF
```{r}
create_params_and_matrices <- function(lam, phi, x_coords, y_coords, sig_kernel) {
  n_points = length(x_coords)
  D <- dist(cbind(x_coords, y_coords), diag = T, upper = T) %>% as.matrix
  W <- exp(-D^2/(2*sig_kernel^2))
  diag(W) = 0
  
  rowsums <- rowSums(W)
  W_tilde <- sweep(W, 1, FUN = '/', STATS = rowsums)
  
  sigma2s <- lam/rowsums

  Q <- diag(1/sigma2s) %*% (diag(n_points) - phi*W_tilde)
  
  return(list(
    'D' = D,
    'W_tilde' = W_tilde,
    'W' = W,
    'rowsums' = rowsums,
    'phi' = phi,
    'Q' = Q
  ))
  
}

simulate_GMRF = function(precision_mat, nsim = 1, mu = rep(0, nrow(precision_mat))) {
  Q = precision_mat
  N = nrow(Q)
  Sigma = solve(Q) 
  out = MASS::mvrnorm(n = nsim, mu = mu, Sigma = Sigma)
  return(out)
}
```

Points simulated in a grid, uniform at random
```{r}
set.seed(123)
N = 20
m = 1000
x_coords <- runif(N)
y_coords <- runif(N)
```


```{r}
sig_kernel = 0.2
phi = 0.95
lam = 1

params_matrices <- create_params_and_matrices(lam, phi, x_coords, y_coords, sig_kernel)
Q <- params_matrices$Q
V <- simulate_GMRF(Q, nsim = m)


phi_true <- params_matrices$phi
W <- params_matrices$W
W_tilde <- params_matrices$W_tilde
lambdas <- eigen(W_tilde)$values
rowsums <- params_matrices$rowsums

```


Discretized distances:
```{r}
D = params_matrices$D

D_vec <- as.vector(D)

n_breaks <- 7
breaks <- quantile(D_vec, probs = seq(0, 1, length.out = n_breaks))


labels_vec <- cut(D_vec, breaks = breaks, include.lowest = T, labels = 1:(n_breaks-1))
labels_numeric <- as.integer(labels_vec)

labels_matrix <- matrix(labels_numeric, nrow = nrow(D), ncol = ncol(D))
diag(labels_matrix) = 0
```

Covariance at distance k and variance (for all diagonal entries)
```{r}
mean_est = mean(V)

cov_at_dist_k_single_v = function(v_vec, S_k_unique, mean_est) {
  summed_vals = apply(S_k_unique, 1, function(idx) {
    i = idx[1]
    j = idx[2]
    v_i = v_vec[i] 
    v_j = v_vec[j]
    (v_i-mean_est)*(v_j-mean_est)
  }) %>% sum
  return(1/nrow(S_k_unique) * summed_vals)
}

cov_at_dist_k <- function(V, k, distance_labels_matrix) {
  if (k == 0) {
    #var_est = apply(V, 2, var) %>% mean
    var_est = var(as.vector(V))
    return(var_est)
  }
  
  m = nrow(V)
  S_k = which(distance_labels_matrix == k, arr.ind = T)
  S_k_unique = S_k[S_k[, "row"] < S_k[, "col"], ]
  
  summed_over_v_vecs = sum(apply(V, 1, function(v_) {
    cov_at_dist_k_single_v(v_, S_k_unique, mean_est)
  }))
  return(1/m*summed_over_v_vecs)
}

```

```{r}
unique_labels <- 0:(n_breaks-1)
cov_by_label <- sapply(unique_labels, function(k) cov_at_dist_k(V, k, labels_matrix))
names(cov_by_label) <- unique_labels


Sigma_hat = matrix(NA, nrow = N, ncol = N)
for (k in unique_labels) {
  Sigma_hat[labels_matrix == k] <- cov_by_label[as.character(k)]
}
Q_est = solve(Sigma_hat)
```




Averaging partial correlations from discretized distances

```{r}
p <- n_breaks-1

beta_est = numeric(p)
beta_est_median = numeric(p)
beta_est_std = numeric(p)
cond_corr_theo = numeric(p)
cond_corr_theo_std = numeric(p)


for (k in 1:p) {
  indices_bin_k <- which(labels_matrix == k, arr.ind = TRUE)
  unique_indices_bin_k <- indices_bin_k[indices_bin_k[,1] < indices_bin_k[,2], ]
  rho_vals_k <- apply(unique_indices_bin_k, 1, function(idx) {
    i <- idx[1]
    j <- idx[2]
    - Q_est[i, j] / sqrt(Q_est[i, i] * Q_est[j, j])
  })
  beta_est[k] <- -mean(rho_vals_k)
  beta_est_median[k] <- -median(rho_vals_k)
  beta_est_std[k] <- sd(rho_vals_k)
  
  # theoretical values
  corr_vals_k <-  apply(unique_indices_bin_k, 1, function(idx) {
    i <- idx[1]
    j <- idx[2]
    phi*W[i, j] / sqrt(rowsums[i] * rowsums[j])
  })
  cond_corr_theo[k] <- mean(corr_vals_k)
  cond_corr_theo_std[k] <- sd(corr_vals_k)
}
beta_est
```


```{r}

# For each bin label, calculate the median distance of all corresponding (unique) pairs.
medians_by_bin <- sapply(1:(n_breaks-1), function(k) {
  idx <- which(labels_matrix == k, arr.ind = TRUE)
  idx_unique <- idx[idx[, "row"] < idx[, "col"], ]
  
  d_vals <- D[idx_unique]
  median(d_vals)
})

```

Plot of scaled gaussian kernel and corresponding betas from discretized distances:

```{r}
gauss_kernel <- function(x) return(exp(-(x^2)/(2*sig_kernel^2)))

avg_row_sum <- mean(rowsums)
mean_kappa_est = mean(diag(Q_est))

# scaling like with the kernel-model on the circle, but here using avg. row sum
scaled_gauss_kernel <- function(x) (return((phi/avg_row_sum)*gauss_kernel(x)))

midpoints <- medians_by_bin

plot(midpoints, -beta_est, pch = 2, ylim = c(min(-beta_est) - 0.025, max(-beta_est) + 0.12), 
     xlab = 'Distance',
     ylab = '',
     lwd = 1.5)

points(midpoints, cond_corr_theo, pch = 4, col = 'red', lwd = 1.5)

arrows(midpoints, -beta_est - beta_est_std, 
       midpoints, -beta_est + beta_est_std,
       angle = 90, code = 3, length = 0.05, col = "black", lty = 'dashed', lwd = 1)



curve(scaled_gauss_kernel, add = TRUE, from = midpoints[1], to = midpoints[p], 
      col ='blue', lwd = 1.5)



legend('topright', pch = c(2, NA, 4), 
       col = c('black', 'blue', 'red'),
       lty = c(NA, 1, NA),
       lwd = c(1.5, 1.5, 1.5),
       legend = c('Estimated betas', 'Scaled kernel', 'Avg. from underlying kernel'),
       cex = 0.8,
       bty = 'n')
title(paste0('N = ', N, ', m = ', m), adj =0, cex.main = 1, line = 0.2)
```




### Plot with varying N and m
```{r}
set.seed(4)

Ns <- c(15, 20, 25, 30)
ms <- c(1000)
combs <- expand_grid(Ns, ms)


###
# define a 3×2 layout matrix 
# cells 1–4 are four panels
# cell 5 (shown twice) spans both columns on the bottom row for the legend.
layout(
  mat = matrix(c(1,2,
                 3,4,
                 5,5),
                  nrow = 3,
                  byrow = TRUE),
  heights = c(1, 1, 0.2)
)

par(mar = c(4, 4, 2, 1))



sig_kernel = 0.2
phi = 0.95
lam = 1

for (i in 1:nrow(combs)) {
  N <- combs$Ns[i]
  m <- combs$ms[i]
  
  ## points 
  x_coords <- runif(N)
  y_coords <- runif(N)
  
  ## matrices
  params_matrices <- create_params_and_matrices(lam, phi, x_coords, y_coords, sig_kernel)
  Q <- params_matrices$Q
  V <- simulate_GMRF(Q, nsim = m)
  
  phi_true <- params_matrices$phi
  W <- params_matrices$W
  W_tilde <- params_matrices$W_tilde
  lambdas <- eigen(W_tilde)$values
  rowsums <- params_matrices$rowsums
  
  ## Distances
  D = params_matrices$D
  D_vec <- as.vector(D)
  
  n_breaks <- 7
  breaks <- quantile(D_vec, probs = seq(0, 1, length.out = n_breaks))
  
  labels_vec <- cut(D_vec, breaks = breaks, include.lowest = T, labels = 1:(n_breaks-1))
  labels_numeric <- as.integer(labels_vec)
  
  labels_matrix <- matrix(labels_numeric, nrow = nrow(D), ncol = ncol(D))
  diag(labels_matrix) = 0
  
  
  ## Precision matrix estimate
  unique_labels <- 0:(n_breaks-1)
  cov_by_label <- sapply(unique_labels, function(k) cov_at_dist_k(V, k, labels_matrix))
  names(cov_by_label) <- unique_labels
  
  Sigma_hat = matrix(NA, nrow = N, ncol = N)
  for (k in unique_labels) {
    Sigma_hat[labels_matrix == k] <- cov_by_label[as.character(k)]
  }
  Q_est = solve(Sigma_hat)
  
  ## Beta estimates
  p <- n_breaks-1

  beta_est = numeric(p)
  beta_est_median = numeric(p)
  beta_est_std = numeric(p)
  cond_corr_theo = numeric(p)
  cond_corr_theo_std = numeric(p)
  
  
  for (k in 1:p) {
    indices_bin_k <- which(labels_matrix == k, arr.ind = TRUE)
    unique_indices_bin_k <- indices_bin_k[indices_bin_k[,1] < indices_bin_k[,2], ]
    rho_vals_k <- apply(unique_indices_bin_k, 1, function(idx) {
      i <- idx[1]
      j <- idx[2]
      - Q_est[i, j] / sqrt(Q_est[i, i] * Q_est[j, j])
    })
    beta_est[k] <- -mean(rho_vals_k)
    beta_est_median[k] <- -median(rho_vals_k)
    beta_est_std[k] <- sd(rho_vals_k)
    
    # theoretical values
    corr_vals_k <-  apply(unique_indices_bin_k, 1, function(idx) {
      i <- idx[1]
      j <- idx[2]
      phi*W[i, j] / sqrt(rowsums[i] * rowsums[j])
    })
    cond_corr_theo[k] <- mean(corr_vals_k)
    cond_corr_theo_std[k] <- sd(corr_vals_k)
  }
  
  ## bin midpoints
  medians_by_bin <- sapply(1:(n_breaks-1), function(k) {
    idx <- which(labels_matrix == k, arr.ind = TRUE)
    idx_unique <- idx[idx[, "row"] < idx[, "col"], ]
    
    d_vals <- D[idx_unique]
    
    median(d_vals)
  })

  gauss_kernel <- function(x) return(exp(-(x^2)/(2*sig_kernel^2)))

  avg_row_sum <- mean(rowsums)
  mean_kappa_est = mean(diag(Q_est))
  
  scaled_gauss_kernel <- function(x) (return(phi/avg_row_sum*gauss_kernel(x)))
  
  
  midpoints <- medians_by_bin
  
  plot(midpoints, -beta_est, pch = 2, ylim = c(min(-beta_est) - 0.08, max(-beta_est) + 0.12), 
       xlab = 'Distance',
       ylab = '',
       lwd = 1.5)
  
  points(midpoints, cond_corr_theo, pch = 4, col = 'red', lwd = 1.5)
  
  arrows(midpoints, -beta_est - beta_est_std, 
         midpoints, -beta_est + beta_est_std,
         angle = 90, code = 3, length = 0.05, col = "black", lty = 'dashed', lwd = 1)
  
  
  curve(scaled_gauss_kernel, add = TRUE, from = midpoints[1], to = midpoints[p], 
        col ='blue', lwd = 1.5)
  title(paste0('N = ', N, ', m = ', m), adj =0, cex.main = 1, line = 0.2)
  
}
# legend seperately
par(mar = c(0, 0, 0, 0))
plot.new()
legend("center",
       legend = c("Estimated betas",
                  "Gaussian kernel (scaled)",
                  "Avg. kernel-based cond. correlations"),
       pch = c(2,   NA,  4),
       lty = c(NA,  1,   NA),
       col = c("black","blue","red"),
       lwd = c(1.5, 1.5, 1.5),
       horiz = TRUE,
       bty = "n",
       cex = 0.9,
       x.intersp = 0.1
)
```

